The core transcriptome of mammalian placentas and the divergence of expression with placental shape

### 2016-11-17 manuscript/manuscript.Rnw
tweak discussion of annexins and other genes

The core placenta transcriptome is significantly enriched in genes
which are involved in multiple biological pathways, including multiple
pathways related to translation (ribosome, translation, signal
recognition particle targeting to the rough endoplasamic reticulum,
etc.) (Supplemental \cref{tab:core_placenta_pathway_analysis}). Beyond
translation, the next most abundant set of significantly enriched
molecular pathways are related to the migration of the trophoblast
into the endometrium, including the EGFR1 pathway and the Actin Y
pathway. Two other pathways, the mPR/Progesterone pathway and Rho cell
motility pathways, feed into the Actin Y pathway as well. Supplemental
\cref{fig:egf_kegg_pathway} shows the EGF pathway in humans with the
median expression in FPKM of genes in the pathway shown. Notably, many
of the genes on the GRB2→MYC pathway are strongly expressed; their
activation leads to an increase in migration and invasion in response
to EGF binding to EGFR1. The Actin cytoskeleton pathway is depicted in
Supplemental \cref{fig:kegg_actin}, which shows that \textit{FN1} and
\textit{ITG} are highly expressed, which activate Rac, and potentially
lead to the formation of focal adhesions and lamellipodia formation
through their inhibition of GSN and HSPC300.

@@

Supplemental \cref{tab:shape_genes} shows the top 50 genes which are
associated with the shape of placenta attachment, as well as the shape
in which the difference is most significant. The most significant
genes which were associated with diffuse placenta attachment include
\textit{EREG}, a ligand of EGFR, \textit{UNC93A}, a homolog of
\textit{UNC93B1} which is involved in the balancing between Tlr7 and
Tlr9 immune response in
mice~\cite{Brinkmann.ea2007:interactionbetweenERmembrane}, and
\textit{WNT5A}, which inhibits the invasion of the trophoblast cell
line HTR8/SVneo \textit{in vitro} \DLA{Cite this paper after reading
  the comments on it (PMID: 26865089)}.

@@

\subsubsection{Immune Regulation}
The transcriptomes from multiple different species enabled us to
identify the set of expressed genes which are common to all placentas.
Components of the annexin complexes, including
\textit{ANXA2}, \textit{ANXA1}, \textit{S100A11}, and \textit{S100A10}
are expressed in all placentas examined, and argue for the evolution
importance of annexins in placenta function, where they likely
establish maternal-fetal
tolerance\cite{Liu.ea2015:AnnexinA2complexeswith,Hutchinson.ea2011:Molecularregulatorsofresolution,Aarli.ea1997:Suppressiveeffectonlymphoproliferation}.
We suggest that further research into molecular role of these annexin
complexes may illuminate how immune tolerance is maintained during
placentation. The core placenta transcriptome is also enriched in
genes involved in maintaining the structure of the placenta and
transporting proteins, as would be expected given the function of the
placenta.

@@

While Syncitin-1 and -2 are implicated in the formation of the
syncytiotrophoblast in
\human~\cite{Rawn.ea2008:evolutionregulationandfunction} and homologs
of \textit{ERVFRD-1} and \textit{ERVW-1} are widely expressed in most
of the placentas analyzed here, they are not expressed in \bonobo,
\sheep, and \dog, even though these mammals form
syncytiotrophoblasts~\cite{Mess.ea2006:Evolutionarytransformationsoffetal}.
Because the formation of syncytiotrophoblast is the stem state of
Eutheria, Afrotheria, Euarchontoglires, and
Laurasiatheria~\cite{Mess.ea2006:Evolutionarytransformationsoffetal},
the gene(s) responsible for cytotrophoblast fusion to form
syncytiotrophoblast is likely to be highly expressed in the core
placenta. \textit{ANXA5} is known to be expressed at multiple stages
of placentation, and is capable of repairing membrane defects and
promoting the fusion of cells~\DLA{Cite more papers here which define
  this.} in the presence of Ca$^{2+}$. While there are likely other
genes which are responsible for the positioning of cytotrophoblast
cells in close proximity to start the fusion process, \textit{ANXA5}
is likely to be essential for the completion of fusion.

\DLA{Possibility of membrane repair due to maternal immune response as
  well during this process, as well as repair of membrane due to high
  secretion and adsoroption of nutrients/waste in the placenta.}

@@

>>>>

The core placenta transcriptome is significantly enriched in genes
which are involved in multiple biological pathways, including multiple
pathways related to translation (ribosome, translation, signal
recognition particle targeting to the rough endoplasamic reticulum,
etc.) (Supplemental \cref{tab:core_placenta_pathway_analysis}). Beyond
translation, the next most abundant set of significantly enriched
molecular pathways are related to the migration of the trophoblast
into the endometrium, including the EGFR1 pathway and the Actin Y
pathway. Two other significant pathways, the mPR/Progesterone pathway
and Rho cell motility pathways, feed into the Actin Y pathway as well.
Supplemental \cref{fig:egf_kegg_pathway} shows the EGF pathway in
humans with the median expression in FPKM of genes in the pathway
shown. Notably, many of the genes on the GRB2→MYC pathway are strongly
expressed; their activation leads to an increase in migration and
invasion in response to EGF binding to
EGFR1~\cite{Gupta.ea2016:CellSignalingPathwaysInvolved}. The Actin
cytoskeleton pathway is depicted in Supplemental
\cref{fig:kegg_actin}, which shows that \textit{FN1} and \textit{ITG}
are highly expressed, which activate Rac, and potentially lead to the
formation of focal adhesions and lamellipodia formation through their
inhibition of GSN and HSPC300. Other compoenents of cell-cell
adhesion, including \textit{F11R} which forms tight junctions in
epithelia, proteglycans \textit{SDC4} and \textit{LUM}, and the
laminin \textit{LAMC1}.

@@

Supplemental \cref{tab:shape_genes} shows the top 50 genes which are
associated with the shape of placenta attachment, as well as the shape
in which the difference is most significant. The most significant
genes which were associated with diffuse placenta attachment include
\textit{EREG}, a ligand of EGFR, \textit{UNC93A}, a homolog of
\textit{UNC93B1} which is involved in the balancing between Tlr7 and
Tlr9 immune response in
mice~\cite{Brinkmann.ea2007:interactionbetweenERmembrane}, and
\textit{WNT5A}, which is a member of the Wnt ligand family whose
singaling pathway has been implicated in motility in human trophoblast
cells~\cite{Gupta.ea2016:CellSignalingPathwaysInvolved}.

@@

The core placenta transcriptome is enriched in multiple genes which
are critical for the function of the placenta, including vesicular
transport, immune regulation, cell fusion and membrane repair, and
invasion.

\subsubsection{Immune Regulation and Annexins}
The transcriptomes from multiple different species enabled us to
identify the set of expressed genes which are common to all placentas.
Components of the annexin complexes, including
\textit{ANXA2}, \textit{ANXA1}, \textit{S100A11}, and \textit{S100A10}
are expressed in all placentas examined, and argue for the evolution
importance of annexins in placenta function, where they likely
establish maternal-fetal
tolerance\cite{Liu.ea2015:AnnexinA2complexeswith,Hutchinson.ea2011:Molecularregulatorsofresolution,Aarli.ea1997:Suppressiveeffectonlymphoproliferation}.
We suggest that further research into molecular role of these annexin
complexes may illuminate how immune tolerance is maintained during
placentation. The core placenta transcriptome is also enriched in
genes involved in maintaining the structure of the placenta and
transporting proteins, as would be expected given the function of the
placenta.

@@

While Syncitin-1 and -2 are implicated in the formation of the
syncytiotrophoblast in
\human~\cite{Rawn.ea2008:evolutionregulationandfunction} and homologs
of \textit{ERVFRD-1} and \textit{ERVW-1} are widely expressed in most
of the placentas analyzed here, they are not expressed in \bonobo,
\sheep, and \dog, even though these mammals form
syncytiotrophoblasts~\cite{Mess.ea2006:Evolutionarytransformationsoffetal}.
Because the formation of syncytiotrophoblast is the stem state of
Eutheria, Afrotheria, Euarchontoglires, and
Laurasiatheria~\cite{Mess.ea2006:Evolutionarytransformationsoffetal},
the gene(s) responsible for cytotrophoblast fusion to form
syncytiotrophoblast is likely to be highly expressed in the core
placenta. \textit{ANXA5} is known to be expressed at multiple stages
of placentation, and is capable of repairing membrane defects and
promoting the fusion of
cells~\cite{Carmeille.ea2015:Annexin-A5promotesmembraneresealing} in
the presence of Ca$^{2+}$. While there are likely other genes which
are responsible for the positioning of cytotrophoblast cells in close
proximity to start the fusion process, \textit{ANXA5} is likely to be
essential for the completion of fusion and the maintenance of membrane
repair during the massive flux of nutrients and waste in the placenta.

% \DLA{Possibility of membrane repair due to maternal immune response as
%   well during this process, as well as repair of membrane due to high
%   secretion and adsoroption of nutrients/waste in the placenta.}

@@

\subsubsection{Invasion: EGF and CD9 in the Core Placenta}

The high level of expression of components of the EGFR signaling
pathway in the core placenta transcriptome (including \textit{GRB2}, \textit{EGFR}, \textit{MAPK3}, \textit{MAP2K2}, \textit{SHC1}, \textit{AKT3}, \textit{AKT2}, \textit{ERBB3}

% Role of CD9 in invasion; commonality between placentas
